Bayesian latent class model with sparse correlation for etiology estimation

### 2017-01-27 Reports/draft.v1/draft_latentquadexp.tex
update report

\subsection{The Joint Density}
With the specification of likelihood and prior, we can construct the joint density needed for building the MCMC algorithm. Let $G_i$ be an indicator of whether subject $i$ has GS measurements, and define the following three index sets.
\begin{align*}
I_1 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=1 \text{ and } G_i=1 \Big\}\\
I_2 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=1 \text{ and } G_i=0 \Big\}\\
I_3 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=0 \Big\}
\end{align*}
then we can define the joint density of data and parameters as follow by combining all building blocks together:
\begin{align*}
& P(M,\Theta, \eta, \gamma, \delta) \\
= & \prod_{i\in I_1} P(M_i^{GS},M_i^{SS},M_i^{BS} | \Theta, \eta, \gamma, \delta) 
\prod_{i\in I_2} P(M_i^{SS},M_i^{BS} |\Theta, \eta, \gamma, \delta)
\prod_{i\in I_3} P(M_i^{BS} |\gamma, \delta) P(\gamma, \delta)
\end{align*}
\\
From the form of this model, we can see that the model we proposed incorporates data from all the sources with measurements of varying quality, and its likelihood for the latent variable adopts a quadratic exponential model with sparse correlation structure, so we will refer to this model as the Latent Sparse Correlation (LSC) model in the rest of this paper. With the joint density fully specified, the posterior distribution of our LSC model can be sampled by a Metropolis-Hastings (within Gibbs) algorithm, of which the details are described in the appendix.

\newpage
\section{Simulation Studies}
\subsection{Data Simulation}
Three sets of simulation studies are carried out to empirically evaluate the effectiveness of the LSC model under different situations. For all three sets of studies, we assume there are five candidate pathogens and two binary covariates. In each study, $200$ independent data sets are simulated, and in each simulated data set, there are $500$ case subjects and $1000$ control subjects.\\

@@

\begin{itemize}
\item[{\bf I}]
In the first set of studies, the true lung infection status $L$ of case patients are generated by a Quadratic Exponential Model, where the first order canonical parameters are dependent on both covariates with an interaction effect, and the second order parameters are independent of the covariates.
\begin{align*}
\theta^{(1)}_{ik} = \beta_{k0} + \beta_{k1} x_{i1} + \beta_{k2} x_{i2} + \beta_{k3} x_{i1} x_{i2}
\end{align*}

Also, for $K = 5$, there are $10$ second order parameters, we assume that two of them are zero, which represents that two particular pairs of pathogens infect lungs independently from each other; and the rest eight association parameters share the same negative value: $-1.5$, which stands for the competition among pathogens. Then, the BS and SS measurements for case subjects are simulated based on formula (8) and (9) respectively, and the SS measurements for control subjects are simulated based on formula (10) assuming that there is no infection at all in control patients' lungs, where $\text{TPR}^{(SS)} \approx 0.1,  \text{TPR}^{(BS)} \approx 0.6, \text{FPR}^{(BS)} \approx 0.45$. And the actual parameter values used in the simulation process are summarized in table \ref{•} in the Appendix.

\item[{\bf II}]
In the second set of studies, the true lung infection status $L$ of case patients are generated by a Multinomial Model, which is equivalent to the above QE model with all the second order parameters set to negative infinity. And the BS and SS measurements are generated in the same way as they are in study I.

\item[{\bf III}]
In the second set of studies, the true lung infection status of case patients are simulated from the same model as in study I, but the parameters that control the measurement quality are set differently, that is $\text{TPR}^{(SS)} \approx 0.8,  \text{TPR}^{(BS)} \approx 0.9, \text{FPR}^{(BS)} \approx 0.1$. Their actually values are summarized in table \ref{•}.\\
\end{itemize}

\subsection{Model Specifications}
In each of the above situation, $200$ independent data sets are generated. The LSC model is applied to each data set (without using GS measurements) with a series of different prior specifications on $\alpha_2, \lambda_a$ and $\lambda_b$, which represent the experts' prior knowledge on the magnitude of the competitions between pathogens. Within each study, the prior specifications share the same values for hyper-parameters $a_k, b_k, c_k, d_k, e_k$ and $f_k$, $k = 1, 2, \ldots, 5$, which control the prior input on measurement quality. Wu, et al.(2015) \cite{wu2015partially} had discussed the model sensitivity to these hyper-parameters and the partial identifiability issue, which also applies to our method, thus we do not further study the sensitivity issue on these hyper-parameters. Their values are selected according to experts' knowledge on the quality of BS and SS measurements, and the true TPR and FPR values are set to be covered by the prior 95\% credible interval. For all three studies, same values are used for $\sigma_1$, and $\sigma_2$, which are large enough to represent non-informativeness. These hyper-parameter values are listed in table \ref{tab: }. \\

@@

\subsection{Results}
Table \ref{tab: } lists the Bhattachayya coefficients of the LSC model under different prior specifications, and of the two pLCMs in Study I. The same information is visualized in figure \ref{}. As we can see, when the true data generating mechanism allows multi-pathogen infection, the classic pLCM (pLCM-1) performs the worst, the pLCM-2 is the second worst, and the LSC model, across all prior specifications, shows a significant amount of improvement over the two pLCM models. The variations caused by different prior specifications is relatively small comparing to the improvement over pLCM models, especially the single-pathogen model.\\

@@

\begin{center}
\includegraphics[scale=0.45]{DataDescription.png}
\captionof{figure}{{\bf Data Description}: }
\label{data}
\end{center}

@@

\begin{table}[htbp]
\begin{adjustwidth}{-0.5in}{-1in}
\resizebox{0.7\textwidth}{!}{\begin{minipage}{\textwidth}
\caption{Summary of the True Parameter Values}
\label{tab: truepar}

\begin{tabular}{llllllllllllllll} 
\hline 
$\mu_1$ & $\mu_2$ & $\mu_3$ & $\pi_0$ & $\pi_1$ & $\pi_2$ & $\pi_3$ & TPR$^{(SS)}_1$ & TPR$^{(SS)}_2$ & TPR$^{(SS)}_3$ & TPR$^{(BS)}_1$ & TPR$^{(BS)}_2$ & TPR$^{(BS)}_3$ & FPR$^{(BS)}_1$ & FPR$^{(BS)}_2$ & FPR$^{(BS)}_3$ \\
\hline \\
$0.584$ & $0.289$ & $0.378$ & $0.10$ & $0.60$ & $0.25$ & $0.05$ & $0.05$ & $0.10$ & $0.10$ & $0.75$ & $0.80$ & $0.70$ & $0.50$ &  $0.50$ & $0.40$\\ \hline
\end{tabular}
\end{minipage}}
\end{adjustwidth}
\end{table}

>>>>

\subsection{The Joint Density}
With the specification of likelihood and prior, we can construct the joint density needed for building the MCMC algorithm. Let $G_i$ be an indicator of whether subject $i$ has GS measurements, and define the following three index sets.
\begin{align*}
H_1 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=1 \text{ and } G_i=1 \Big\}\\
H_2 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=1 \text{ and } G_i=0 \Big\}\\
H_3 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=0 \Big\}
\end{align*}
then we can define the joint density of data and parameters as follow by combining all building blocks together:
\begin{align*}
& P(M,\Theta, \eta, \gamma, \delta) \\
= & \prod_{i\in H_1} P(M_i^{GS},M_i^{SS},M_i^{BS} | \Theta, \eta, \gamma, \delta) 
\prod_{i\in H_2} P(M_i^{SS},M_i^{BS} |\Theta, \eta, \gamma, \delta)
\prod_{i\in H_3} P(M_i^{BS} |\gamma, \delta) P(\gamma, \delta)
\end{align*}
\\
From the form of this model, we can see that the model we proposed incorporates data from all the sources with measurements of varying quality, and its likelihood for the latent variable adopts a quadratic exponential model with sparse correlation structure, so we will refer to this model as the Latent Sparse Correlation (LSC) model in the rest of this paper. With the joint density fully specified, the posterior distribution of our LSC model can be sampled by a Metropolis-Hastings (within Gibbs) algorithm, of which the details are described in the appendix.

\newpage
\section{Simulation Studies}
\subsection{Data Simulation}
Three sets of simulation studies are carried out to empirically evaluate the effectiveness of the LSC model under different situations. For all three sets of studies, we assume there are five candidate pathogens ($A, B, \ldots, E$) and two binary covariates ($X_1$ and $X_2$). In each study, $200$ independent data sets are simulated, and in each simulated data set, there are $500$ case subjects and $1000$ control subjects.\\

@@

\begin{itemize}
\item[{\bf I}]
In the first set of studies, the true lung infection status $L$ of case patients are generated by a Quadratic Exponential Model, where the first order canonical parameters are dependent on both covariates with an interaction effect, and the second order parameters are independent of the covariates. For case subject $i$ and the $k$th pathogen, 
\begin{align*}
\theta^{(1)}_{ik} = \beta_{k0} + \beta_{k1} x_{i1} + \beta_{k2} x_{i2} + \beta_{k3} x_{i1} x_{i2}
\end{align*}

Also, with $K = 5$, there are $10$ second order parameters, we assume that two of them are zero, which represents that two particular pairs of pathogens ($B:C$ and $B:D$) infect lungs independently from each other; and the rest eight association parameters share the same negative value: $-1.5$, which stands for the pairwise competition among pathogens. Then, the BS and SS measurements for case subjects are simulated based on formula (8) and (9) respectively, and the SS measurements for control subjects are simulated based on formula (10) assuming that there is no infection at all in control patients' lungs, where $\text{TPR}^{(SS)} \approx 0.1,  \text{TPR}^{(BS)} \approx 0.6, \text{FPR}^{(BS)} \approx 0.45$. And the actual parameter values used in the simulation process are summarized in table \ref{tab:study1}.

\begin{table}[h]
\centering
\caption{The model parameters used for data simulation in study I}
\label{tab:study1}
\begin{tabular}{lrrrrr}
\hline
Pathogen     & A    & B     & C     & D     & E    \\ \hline
$\beta_0$    & 0.21 & -0.28 & -0.84 & -0.21 & 1.07 \\
$\beta_1$    & -0.1 & -0.5  & 0.5   & 0.2   & 0.1  \\
$\beta_2$    & -0.3 & 0.2   & -0.2  & -0.1  & 0.3  \\
$\beta_3$    & 0.4  & 0.3   & -0.4  & 0.2   & -0.2 \\
TPR$^{(SS)}$ & 0.11 & 0.12  & 0.08  & 0.15  & 0.10 \\
TPR$^{(BS)}$ & 0.80 & 0.60  & 0.70  & 0.70  & 0.50 \\
FPR$^{(BS)}$ & 0.50 & 0.55  & 0.40  & 0.35  & 0.45 \\ \hline
\end{tabular}
\end{table}

\item[{\bf II}]
In the second set of studies, the true lung infection status $L$ of case patients are generated by a Multinomial Model, which is equivalent to the above QE model but with all the second order parameters set to negative infinity. The multinomial etiology probabilities are listed in table \ref{tab:study2}. And the BS and SS measurements are generated in the same way as they are in study I with the same TPRs and FPRs.

\begin{table}[h]
\centering
\caption{The etiology probabilities used for data simulation in study II}
\label{tab:study2}
\begin{tabular}{lrrrrrr}
\hline
         & Other & A     & B     & C     & D     & E     \\ \hline
strata 1 & 0.200 & 0.241 & 0.079 & 0.088 & 0.125 & 0.267 \\
strata 2 & 0.171 & 0.224 & 0.113 & 0.106 & 0.191 & 0.196 \\
strata 3 & 0.232 & 0.191 & 0.049 & 0.115 & 0.105 & 0.308 \\
strata 4 & 0.224 & 0.176 & 0.072 & 0.137 & 0.162 & 0.230 \\ \hline
\end{tabular}
\end{table}

\item[{\bf III}]
In the third set of studies, the true lung infection status of case patients are simulated from the same model as in study I, but the parameters that control the measurement quality are set differently, that is $\text{TPR}^{(SS)} \approx 0.8,  \text{TPR}^{(BS)} \approx 0.9, \text{FPR}^{(BS)} \approx 0.05$. Their values are listed in table \ref{tab:study3}.\\

\begin{table}[h]
\centering
\caption{The model parameters used for data simulation in study III that are different from study I}
\label{tab:study3}
\begin{tabular}{lrrrrr}
\hline
Pathogen     & A    & B     & C     & D     & E    \\ \hline
TPR$^{(SS)}$ & 0.81 & 0.82  & 0.88  & 0.85  & 0.80 \\
TPR$^{(BS)}$ & 0.98 & 0.96  & 0.97  & 0.97  & 0.95 \\
FPR$^{(BS)}$ & 0.050 & 0.055  & 0.040  & 0.035  & 0.045 \\ \hline
\end{tabular}
\end{table}

\end{itemize}

\subsection{Model Specifications}
In each of the above situation, $200$ independent data sets are generated. The LSC model is applied to each data set (without using GS measurements) with a series of different prior specifications on $\alpha_2, \lambda_a$ and $\lambda_b$, which represent the experts' prior knowledge on the magnitude of the competitions between pathogens. Within each study, the values for hyper-parameters $a_k, b_k, c_k, d_k, e_k$ and $f_k$, $k = 1, 2, \ldots, 5$, which control the prior input on measurement quality, do not vary. Wu, et al.(2015) \cite{wu2015partially} had discussed the model sensitivity to these hyper-parameters and the partial identifiability issue, which also applies to our method, thus we do not further study the sensitivity issue on these hyper-parameters. Their values are selected according to experts' knowledge on the quality of BS and SS measurements, and the true TPR and FPR values are set to be covered by the prior 95\% credible interval. For all three studies, same values are used for $\sigma_1$, and $\sigma_2$, which are set large enough to represent non-informativeness. These hyper-parameter values are listed in table \ref{tab:hyper}. \\

\begin{table}[h]
\centering
\caption{The common hyper-parameters used for model fitting in simulation studies}
\label{tab:study3}
\begin{tabular}{lrrrrrrrr}
\hline
          & $a_k$    & $b_k$     & $c_k$     & $d_k$     & $e_k$ & $f_k$ & $\sigma_1$  & $\sigma_2$    \\ \hline
Study I   & 7.6 & 59  & 5  & 3  & 1 & 1 & 0.2 & 0.2 \\
Study II  & 50 & 10  & 12  & 1  & 1 & 12 & 0.2 & 0.2 \\
Study III & 7.6 & 59  & 5  & 3  & 1 & 1 & 0.2 & 0.2 \\ \hline
\end{tabular}
\end{table}

@@

\subsection{Results}
Table \ref{tab:result1} lists the Bhattachayya coefficients of the LSC model under different prior specifications, and of the two pLCMs in Study I. The same information is visualized in figure \ref{fig:study1}. As we can see, when the true data generating mechanism allows multi-pathogen infection, the classic pLCM (pLCM-1) performs the worst, the pLCM-2 is the second worst, and the LSC model, across all prior specifications, shows a significant amount of improvement over the two pLCM models. The variations caused by different prior specifications is relatively small comparing to the improvement over pLCM models, especially the single-pathogen model.\\

\begin{table}[h]
\centering
\caption{Summary of the overall parameter estimation accuracy of each model fitted in study I}
\label{tab:result1}
\begin{tabular}{llll}
\hline
$\alpha_2$ & $\lambda_a$ & $\lambda_b$ & Bhattacharyya \\ \hline
-5         & 6           & 2           & 0.9838        \\
-3.5       & 6           & 2           & 0.9828        \\
-7.5       & 4           & 4           & 0.9828        \\
-5         & 4           & 4           & 0.9815        \\
-2         & 6           & 2           & 0.9790         \\
-2         & 15          & 5           & 0.9790         \\
-1.5       & 6           & 2           & 0.9770         \\
-3.5       & 4           & 4           & 0.9765        \\
-1         & 6           & 2           & 0.9748        \\
-7.5       & 15          & 5           & 0.9747        \\
-10        & 4           & 4           & 0.9742        \\
-3.5       & 15          & 5           & 0.9728        \\
-2         & 2           & 2           & 0.9712        \\
-7.5       & 6           & 2           & 0.9711        \\
-1         & 15          & 5           & 0.9677        \\
-1         & 2           & 2           & 0.9673        \\
-10        & 6           & 2           & 0.9649        \\
-10        & 15          & 5           & 0.9604        \\ \hline
\multicolumn{3}{l}{pLCM-2}             & 0.9494        \\
\multicolumn{3}{l}{pLCM-1}             & 0.7681        \\ \hline
\end{tabular}
\end{table}

\begin{center}
\includegraphics[scale=0.6]{StudyI.png}
\captionof{figure}{{\bf Overall estimation accuracy of models in study I}: }
\label{fig:study1}
\end{center}

@@

\begin{center}
\includegraphics[scale=0.45]{DataDescription.png}
\captionof{figure}{{\bf Data Description}: }
\label{}
\end{center}

@@

%\begin{table}[htbp]
%\begin{adjustwidth}{-0.5in}{-1in}
%\resizebox{0.7\textwidth}{!}{\begin{minipage}{\textwidth}
%\caption{Summary of the True Parameter Values}
%\label{tab: truepar}

%\begin{tabular}{llllllllllllllll} 
%\hline 
%$\mu_1$ & $\mu_2$ & $\mu_3$ & $\pi_0$ & $\pi_1$ & $\pi_2$ & $\pi_3$ & TPR$^{(SS)}_1$ & TPR$^{(SS)}_2$ & TPR$^{(SS)}_3$ & TPR$^{(BS)}_1$ & TPR$^{(BS)}_2$ & TPR$^{(BS)}_3$ & FPR$^{(BS)}_1$ & FPR$^{(BS)}_2$ & FPR$^{(BS)}_3$ \\
%\hline \\
%$0.584$ & $0.289$ & $0.378$ & $0.10$ & $0.60$ & $0.25$ & $0.05$ & $0.05$ & $0.10$ & $0.10$ & $0.75$ & $0.80$ & $0.70$ & $0.50$ &  $0.50$ & $0.40$\\ \hline
%\end{tabular}
%\end{minipage}}
%\end{adjustwidth}
%\end{table}
